\chapter{Environments}
\label{chap-environments}

Recall that an \emph{environment} is a mapping from \emph{names} to
\emph{objects}.  This mapping consists of a set of \emph{bindings}. 

When a user is created in the system, a \emph{default global
  environment} is created for that user.   The global environment of a
user consists of a \emph{system-wide} environment and a
\emph{user-specific} environment.

The system-wide environment consists of bindings that are themselves
immutable (i.e., the user is not allowed to alter the binding) such as
the binding of the symbol \texttt{cl:length} to the function that
returns the length of a sequence.%
\footnote{It may be necessary to allow the user to change bindings
  such as the one of \texttt{cl:length} to the function that returns
  the length of a sequence.  In fact, it may be necessary to allow the
  user to modify every binding, in which case the global environment
  for a user contains no \emph{system-wide} environment.
  Alternatively, the system-wide environment would be reduced to a
  small set of bindings.  Perhaps bindings that allow the user to
  recover after destroying his or her environment should be stored
  there. }
The objects of these bindings are
also immutable, such as the length function itself.  The system-wide
environment is the same for every user, allowing the installation of
software that is immediately visible to all users. 

The user-specific environment consists of bindings that are created by
the user.  These bindings are of three different kinds:

\begin{itemize}
\item Bindings created by the user for instance as a result of
  executing a \texttt{defparameter} or \texttt{defun} form.
\item Default system-wide bindings that can be altered by the user,
  such as the value of \texttt{*print-base*}.
\item Immutable bindings where the \emph{object} can be modified by
  the user, such as system-defined generic functions to which the user
  is allowed to add specific methods.  Each user has a private copy of
  such objects.
\end{itemize}

The environment contains the following mappings:%
\footnote{I may have forgotten some mappings that are part of the
  global environment.}


\begin{itemize}
\item Mappings from names to \emph{packages} as managed by
  \texttt{make-package}, \texttt{delete-package}, \texttt{defpackage},
  etc. 
\item Mappings from names to \emph{function objects}, as managed by
  \texttt{symbol-function}, \texttt{(setf symbol-function)},
  \texttt{fdefinition}, and \texttt{(setf fdefinition)}.
\item Mappings from names to \emph{macro functions}, as managed by 
\texttt{macro-function} and \texttt{(setf macro-function)}.
\item Mappings from names to compiler macros, as managed by
  \texttt{compiler-macro-function} and \texttt{(setf
    compiler-macro-function)}. 
\item Mappings from names to \emph{classes} as managed by
  \texttt{find-class}, \texttt{(setf find-class)}, \texttt{defclass},
  \texttt{defstruct}, \texttt{define-condition}, etc. 
\item Mappings from names to \emph{type definitions} established by
  \texttt{deftype}. 
\item Mappings from names to \emph{global symbol macros} defined by
  \texttt{define-symbol-macro}.
\item Mappings from names to \emph{constant variables} defined by
  \texttt{defconstant}.
\item Mappings from names to \emph{special variables} defined by
  \texttt{(proclaim special)}, \texttt{defvar}, etc. 
\end{itemize}

When a function or method object is created as a result of calling
\texttt{compile} on a lambda expression, or as a result of loading a
\emph{fasl} file, the object is \emph{linked} to the current global
environment, in that external references are then resolved.  When such
a function or method object is given to a different user, that
different user can execute it, but external references in it will
still refer to the environment into which it was compiled or loaded.

Notice that methods are not in themselves part of the environment.
When we say that a method is \emph{linked} to the current global
environment, we just mean that references to symbols within that
method are resolved in the current global environment.  

This mechanism provides an efficient method of protection.  A user A
can grant controlled access to part of his or her global environment
by allowing a user B to execute a function made available to him or
her through the \emph{object store}.  \seechap{chap-object-store} In a
traditional modern operating system such as \unix{}, this kind of
controlled access required the use of the \emph{setuid} mechanism,
simply because in such a system there is no way to access an object
other than through the global file system, and the accessing user must
have the right permissions to access the object.

The same mechanism can be used by the system itself to protect objects
that would be unwise to give users direct access to, such as disks or
printers. 

