\chapter{Protection}
\label{chap-protection}

There are two kinds of protection that are important in an operating
system:

\begin{itemize}
\item \emph{protecting different users from each other}.  User A
  should not be able to access or destroy the data of some other user
  B, other than if B explicitly permits it, and then only in ways that
  are acceptable to B.
\item \emph{protecting the system from the users}.  Users should be
  able to access system resources such as memory and peripherals only
  in controlled ways, so as to guarantee the integrity of the system.
\end{itemize}

\section{Protecting users from each other}

We use a combination of \emph{access control lists} and
\emph{capabilities}.  All heap-allocated objects except \texttt{cons}
cells and (heap-allocated) numbers are manipulated through a
\emph{tagged pointer}.  In addition to containing a type tag, the
pointer also contains an \emph{access tag}.  The access tag consists
of the 4 most-significant bits of a 64-bit pointer.  Before a pointer
is used to fetch an object from memory, the access bits are cleared.
A primitive operation to fetch the access tag of a pointer is
available to any user code.  Each of the 4 bits represents a potential
\emph{access restriction}, the significance of which is up to the
programmer.  A function that wishes to restrict permission to some
object can test the corresponding access bit and signal an error if
that bit is set.

The author of some complex data structure may for instance grant
access to it only to certain other users.  This would be done by
interpreting one of the access bits as \emph{read permission}, and by
having generic functions that access the data structures check that
this bit has the desired value (for instance in a \texttt{:before}
method). 

The access bits of a capability are determined when the object is
accessed through the object store.  \seechap{chap-object-store} One of
the possible attributes associated with the object in the object store
corresponds to the access permissions in the form of an \emph{access
  control list}.  A user who accesses the object from the object store
will be checked against the access control list and appropriate access
bits will be cleared in the object before it is given to the user.

\section{Protecting the system from the users}

In a typical modern operating system, the system is protected from the
users through the use of a \emph{mode} of execution of the processor,
which can be either \emph{user mode} or \emph{supervisor mode}.
Certain instructions are restricted to supervisor mode, such as
instructions for input/output or for remapping the address space. 

In \sysname{}, the normal mode of execution is \emph{supervisor mode}.
The code executed by the user is translated to machine code by a
\emph{trusted compiler} which is known not to generate code that, if
executed, might represent a risk to the integrity of the system.
Since no remapping of the address space is required as a result of an
\emph{interrupt} or a \emph{trap}, such events can be handled very
quickly. 

Occasionally, it might be useful to write or install some software
that is compiled to machine code by some compiler that can not be
trusted.  The result of such a compilation or installation is a single
(possibly large) Lisp function.  When this function is executed, the
mode of execution is switched to \emph{user mode}.  As with
traditional modern operating systems, the code of such software has
its own \emph{address space}, which means that it can not directly
manipulate \sysname{} capabilities.  Instead, it has to communicate
with the system through the user of \emph{system calls}.  A
system-wide object is referred to by such code through an interposing
\emph{object descriptor}, much like a file descriptor in \unix{}.  The
details of this mechanism have not yet been fully determined.


